---
title: "PA1_template.Rmd"
author: "Donough Regan"
date: "25 January 2016"
output: html_document
---
This is my solution to Course Project 1 of the Reproducible Research Module described
at https://www.coursera.org/learn/reproducible-research/peer/gYyPt/course-project-1

- **Loading and preprocessing the data**

```{r}
echo=TRUE
activity<-read.csv("activity.csv")
names(activity)
#note that missing values are coded as NA
head(activity)
```
- **What is mean total number of steps taken per day?**
```{r}
## Mean total number of steps per day. Ignore missing values.
echo=TRUE
steps_per_day<-tapply(activity$steps,activity$date,sum,na.rm=T)
hist(steps_per_day,xlab="total steps per day",main="Histogram of steps per day")
```
```{r}
echo = TRUE
# Median and mean number of steps per day
median_steps<-median(steps_per_day)
mean_steps<-mean(steps_per_day)
c(median_steps,mean_steps)
```

- **What is the average daily activity pattern?**
```{r}
echo=TRUE
mean_steps_per_interval<-tapply(activity$steps,activity$interval,mean,na.rm=T)
df_activity<-data.frame(interval=unique(activity$interval),steps=mean_steps_per_interval)
plot(df_activity,type="l",xlab="5 minute interval",ylab="Mean steps per interval",main="Average steps made in each 5 minute interval")
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
echo=TRUE
mean_steps_per_interval[which.max(mean_steps_per_interval)]
```

- **Imputing missing values**
Strategry: Create a function which will go through the values when it gets to a NA value then use the mean number of steps for the corresponding interval

```{r}
# Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
echo=TRUE
sum(is.na(activity$steps))
# Fill in the missing data
activity_new<-activity
for(i in 1:nrow(activity)){
  if(is.na(activity$steps[i])){
    replace_index<-activity$interval[i]
    activity_new$steps[i]=mean_steps_per_interval[[as.character(replace_index)]]
  }
}
```

```{r}
echo = TRUE
# Now for the histogram of total steps per day
steps_per_day_new<-tapply(activity_new$steps,activity_new$date,sum,na.rm=T)
hist(steps_per_day_new,xlab="total steps per day",main="Histogram of steps per day")
# Mean and median
median_steps_new<-median(steps_per_day_new)
mean_steps_new<-mean(steps_per_day_new)
c(median_steps,median_steps_new)
c(mean_steps,mean_steps_new)
```
Impact of imputing missing data on estimates of the total daily number of steps
```{r}
echo = TRUE
summary(activity)
summary(activity_new)
```
Conclusion: properties look the same aside from the fact that in activity_new there are no NA values.

- **Are there differences in activity patterns between weekdays and weekends?**

I've two approaches and I'll list them both below.

    - Approach 1: Create variable according to weekday/weekend. Then filter/subset the data. Use the subsetted data to get the mean number of steps for the weekday/weekend. Combine them in a dataframe and make the plot separated by category of weekday/weekend.
```{r}
echo=TRUE
activity_new$weekday_end<-c("weekday")
activity_new[weekdays(as.Date(activity_new$date)) %in% c("Saturday","Sunday"),]$weekday_end<-c("weekend")
activity_new$weekday_end<-factor(activity_new$weekday)
table(activity_new$weekday_end)
activity_new_weekday<-subset(activity_new,weekday_end=="weekday")
activity_new_weekend<-subset(activity_new,weekday_end=="weekend")
meansteps_weekday<- tapply(activity_new_weekday$steps,activity_new_weekday$interval,mean)
meansteps_weekend<- tapply(activity_new_weekend$steps,activity_new_weekend$interval,mean)
df_activity_weekday<-data.frame(interval=unique(activity_new_weekday$interval),avgsteps=meansteps_weekday,weekday_end="weekday")
df_activity_weekend<-data.frame(interval=unique(activity_new_weekend$interval),avgsteps=meansteps_weekend,weekday_end="weekend")
df_activity<-rbind(df_activity_weekday,df_activity_weekend)
library(lattice)
xyplot(avgsteps~interval | weekday_end, data=df_activity,layout=c(1,2),type="l",xlab="Interval",ylab="Number of steps")
```

    - Approach 2: Thought the above approach was a little cumbersome. A more elegant solution is to use the dplyr package.
```{r}
library (dplyr)
echo=TRUE
steps_per_interval_weekend<-activity_new %>% filter(weekday_end %in% c("weekend")) %>% group_by(interval) %>% summarise(avgsteps=mean(steps),weekday_end=factor("weekend"))
steps_per_interval_weekday<-activity_new %>% filter(weekday_end %in% c("weekday")) %>% group_by(interval) %>% summarise(avgsteps=mean(steps),weekday_end=factor("weekday"))
steps_per_interval<-rbind(steps_per_interval_weekend,steps_per_interval_weekday)
#xyplot(avgsteps~interval | weekday_end,layout=c(1,2),data=steps_per_interval,type="l",xlab="Interval",ylab="Number of steps")
```
